Up to 25\% of beneficial mutations in protein sequences are not adaptive innovations in mammals

### 2022-11-04 manuscript/main.tex
References to figures and tables

    Adaptation is considered as the main process driving the diversity of forms and functions across the tree of life, as it enables species to access new ecological niches and also respond to changes in their environment\cite{darwin_origin_1859}.
    In this context of adaptative evolution, a change in environment is echoed as a change of species' forms and functions, ultimately leaving signatures of accelerated changes in the species' genome\cite{merrell_adaptive_1994}.
    This signature of an accelerated rate of changes, generating an excess of beneficial mutations is used to quantify adaptation (fig.~\ref{fig:fitness-landscape}A), for specific genes along a specific lineage\cite{mcdonald_adaptative_1991, smith_adaptive_2002a, welch_estimating_2006}.
    The current availability of large scale genomic data and the development of theoretical models allowed to detect and quantify adaption across genes and lineages, which is now a common practice in evolutionary biology\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    Although these approaches led to a better understanding of the processes shaping molecular evolution, it had the effect to merely associate beneficial mutations with adaptive evolution, which should be dissociated since adaptive evolution is not the only process which can lead to beneficial mutations\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    Indeed, even in a constant environment for which the population is already adapted, a deleterious mutation in an individual can reach fixation in the population by genetic drift\cite{Ohta1992}.
    Subsequently, a mutation restoring the ancestral state will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    More precisely, this restoration of the ancestral state can happen at another loci, in which case it is referred as compensating mutations\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    Conversely, the restoration of the ancestral state can happen at the same locus, in which case it is referred as a beneficial back-mutation\cite{piganeau_estimating_2003, charlesworth_other_2007}, which are the focus of this study.
    Taken together, not all beneficial mutations contribute to genetic innovation, and truly adaptive evolution can only be well estimated under the background of beneficial back-mutations\cite{keightley_what_2010, rice_evolutionarily_2015}.
    It is thus essential to distinguish the underlying reason for the advantage brought by a beneficial mutation, a back-mutation repairing damaged genomes or creating an adaptive innovations.

    For an individual carrying a beneficial mutation, whether this mutation is new innovation as response to a change in environment or a beneficial back-mutation is indistinguishable.
    Likewise, at the level of the population, both will result in a positive transmission bias of the beneficial allele.
    However, their consequences at the macro-evolutionary scale are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promotes phenotype stability and preserves well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Fundamentally, adaptive evolution is unpredictable because it caused by an unpredictable change in the environment and the underlying fitness landscape\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known.
    As a result, a way to quantify the amount of back-mutation is to estimate the underlying fitness landscape, assuming it is constant and stable across time.

    In this study, fitness landscapes of protein-coding genes are estimated from multi-species sequence alignments.
    Multi-species sequence alignments can be used as a proxy for fitness, since for example a deleterious amino acid will be more likely to be absent in the alignment.
    One of the most used methods to make these predictions is SIFT (Sort Intolerant From Tolerant), which reliably detects deleterious mutations in many taxa\cite{ng_sift_2003, vaser_sift_2016}.
    However, SIFT scores is not interpretable as a selection coefficient.
    Second, we know that a lot of mechanisms can influence amino acid frequency other than fitness (phylogenetic inertia, mutation biases, GC-biased gene conversion, mutational paths) which are not accounted for in SIFT\@.
    Contrarily, so-called mutation-selection codon models are rooted within a population-genetic formalism\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    These phylogenetic codon models can disentangle processes of mutation at the DNA level and selection at the amino acid level.
    Such codon models provide a nearly-neutral framework of evolution by estimating the fitness landscape over amino acid sequences for each site of the sequence (fig~\ref{fig:method}A) while taking into account non-adaptive processes impacting codon frequency\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022a}.
    From the fitness landscape obtained by fitting the model to the data, one can thus compute the scaled selection coefficient, denoted $\Sphy$, of any non-synonymous mutations.
    Importantly, mutation-selection codon models assume that the underlying fitness landscape is stable along the phylogeny (as in fig.~\ref{fig:fitness-landscape}B), hence the null subscript $0$ for $\Sphy$.
    As a result, a mutation with a positive $\Sphy$ would be driving the current site toward a fitter amino-acid, and thus considered a beneficial back-mutation repairing existing functions.
    Altogether, our expectation if that beneficial back-mutations are detectable and scattered across the genomes, while at the same time some slightly deleterious mutations are reaching fixations at other loci, generating a balance at which genomes are constantly both damaged and repaired at different loci.
    Even though each of these back-mutations is expected to have a small beneficial effect on their bearer, and that our assumptions may not always be valid, we expect that the genome-wide signature of beneficial back-mutations can be detected and quantified.

    By integrating datasets at different scales (phylogeny and population), we can predict whether substitutions in terminal lineages or any mutation currently segregating in a population are beneficial back-mutations, i.e with a positive $\Sphy$ (fig~\ref{fig:method}B).
    Such a dataset can allow us to address different questions regarding the selection affecting protein-coding DNA sequences.
    Are we empirically observing beneficial back-mutations that reached fixation recently, and one that are currently segregating in populations?
    Are they effectively beneficial for the individual carrying it?
    Conversely, what is the probability for a beneficial mutation to be a back-mutation?
    In other words, can we find empirical evidence and can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    To answer these questions, we estimated the fitness landscape of amino-acids for long-term evolution using available multiple sequence alignments between mammalian species\cite{ranwez_orthomam_2007, howe_ensembl_2021}.
    We then integrated data at the phylogenetic and population scale to estimate the amount of beneficial mutations across the entire exome for 28 populations from 6 genera (\textit{Equus},  \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}; fig.~\ref{fig:method}A\&B).

@@

    \subsection*{Back-mutations in the terminal lineages}
    We first probed for empirical evidence of beneficial back-mutation in the terminal lineages, which should be effectively beneficial.
    We thus reconstructed the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the mammals' phylogeny (fig.~\ref{fig:method}A).
    We showed that among all the substitutions in the terminal lineages, between 10 and 13\% had $\Sphy > 1$ (table SX) and between 25 and 30\% had $\Sphy > 0$ (table SX).
    Substitutions with $\Sphy > 1$ have been beneficial since the ratio of non-synonymous over synonymous divergence, called $\dnds$, was estimated to be between 1.15 and 1.68 in the different lineages (table SX).
    This meant that beneficial back-mutations with $\Sphy > 1$ reach fixation more frequently than synonymous mutations, which are supposed to be neutral, thus are effectively beneficial.
    This also means that the ratio $\dnds$ usually used to detect adaptation while relying on substitutions between close species\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is overinflated due to beneficial back-mutations.
    Then, $\dnds$ computed exclusively on non-synonymous sites with a $\Sphy < 0$ only represent the rate of evolution discarding beneficial back-mutations by construction.
    Thus, if we divide the $\dnds$ estimated on all substitutions by the one estimated only on substitutions predicted to be deleterious, we directly get the proportion of the rate of evolution which is contributed by beneficial back-mutations (see methods).
    This way, we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations (table SX).

    \subsection*{Back-mutations in the populations}
    Given that beneficial back-mutations are effectively selected for, we then ask which proportion of new mutations are beneficial back-mutations.
    For each population with available polymorphism data, we first reconstructed the ancestral sequence (instead of the reference sequence) at the base of the population genealogy.
    From this reference sequence, we computed the scaled selection coefficient ($\Sphy$) for each possible non-synonymous mutations.
    We thus obtained the distribution of fitness effects (DFE) of new mutations from the ancestral sequence in \textit{Homo sapiens} (fig.~\ref{fig:homo-afr-results}A), under the assumption of the stable fitness landscape.
    Because $\Sphy$ is based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ are considered beneficial back-mutations.
    Beneficial back-mutations only represents ~1.5\% of all possible new mutations (red in fig.~\ref{fig:homo-afr-results}A, table~SX).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents ~8.5\% of all observed mutations (red in fig.~\ref{fig:homo-afr-results}B, table~SX).

    Additionally, in populations, the frequencies at which SNPs are segregating is informative of their selective effect, e.g. deleterious SNPs are segregating at lower frequencies because they are purified.
    As such, methods have been developed to estimate the distribution of fitness effect (DFE) at the population scale by gathering information across many SNPs\cite{eyre-walker_distribution_2006a, eyre-walker_estimating_2009a, galtier_adaptive_2016, tataru_inference_2017}.
    These methods offer a unique opportunity to compare the DFE at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    In order to estimate the DFE at the population level and gather robust statistical estimates, we grouped SNPs observed at the population level (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}) in 5 classes of predicted selection coefficient: severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).
    In each of theses classes, the number of derived alleles in the population for each SNPs is used to generate a site-frequency spectrum (SFS) (fig.~\ref{fig:homo-afr-results}C).
    Compared to synonymous mutations which are supposedly neutral (black line in fig.~\ref{fig:homo-afr-results}C), mutations predicted to be deleterious (blue and green lines in fig.~\ref{fig:homo-afr-results}C) were, as expected, underrepresented at high frequencies in the population, a signature of ongoing purifying selection.
    However, these observations remain qualitative.

    To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE\cite{tataru_inference_2017, tataru_polydfe_2020}.
    This model uses the count of derived alleles for each frequency class of the SFS to infer the distribution of fitness effects (DFE) of the region under study.
    The model uses the number of sites ($L$) for each category (fig.~\ref{fig:homo-afr-results}A, see Material and Methods), the counts of synonymous derived allele as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral state.
    We could estimate the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogeny of mammals were effectively purged at the population scale.
    Purifying selection is therefore largely predictable and amino acids with low fitness across mammals are effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the scale of the phylogeny of mammals are indeed beneficial for individuals currently bearing them.
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange in fig.~\ref{fig:homo-afr-results}) were effectively composed of neutral, beneficial and deleterious mutations.

    We replicated our experiment across 28 populations of different mammal species (fig.~\ref{fig:heatmap}) and found that SNPs predicted to be deleterious were indeed purged at the population scale (blue and green rows in fig.~\ref{fig:heatmap}A).
    However, SNPs predicted to be weakly beneficial ($\divWeakAdv$) were composed of a mix of neutral and selected mutations with varying proportions across the different populations (orange row in fig.~\ref{fig:heatmap}B).
    Finally, SNPs predicted to be beneficial back-mutations ($\divAdv$) were found to be beneficial with varying proportions across populations (red row in fig.~\ref{fig:heatmap}C).
    The variable proportions found between populations can be explained by the effective number of individuals in the population ($\Ne$), which is well-known to be a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, we used the Watterson $\theta_w$ as a proxy to order populations by their levels of synonymous diversity.
    Higher synonymous diversity is associated with a smaller proportion of neutral mutations (fig.~\ref{fig:diversity}A).
    This result confirms the nearly-neutral theory and suggests that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) are more likely to discriminate between beneficial and deleterious mutations (fig~SX-SX), while mutations in populations with low diversity (e.g.~\textit{Homo}) are effectively nearly-neutral.

@@

    Finally, we propose a new measure to estimate the probability for a beneficial mutation to be a back-mutation, i.e. $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$.
    We obtained from the analyses above the following probabilities $\mathbb{P} [ \Sphy > 1 ]$, $\mathbb{P} [ \Spop > 1 ]$, $\mathbb{P} [ \Spop > 1  \given  \Sphy > 1]$.
    First, $\mathbb{P} [ \Sphy > 1 ]$ is the probability for a beneficial back-mutation ($\Sphy > 1$) from the ancestral sequence (red fill of histogram in fig~\ref{fig:homo-afr-results}A and table S2).
    Second, $\mathbb{P} [ \Spop > 1 ]$ is the probability of mutation to be beneficial at the level of the population ($\Spop > 1$), estimated from all available SNPs (black fill in fig~\ref{fig:diversity}A and table S2).
    Finally, $\mathbb{P} [ \Spop > 1  \given  \Sphy > 1]$ is the probability for a back-mutation ($\Sphy > 1$) to be effectively beneficial at the level of the population ($\Spop > 1$), when we restricted our analysis on SNPs that are already predicted to be beneficial back-mutations (black fill for the category $\Sphy > 1$ in fig~\ref{fig:homo-afr-results}D and table S2).
    Combining these probabilities using Bayes theorem, we now obtain the probability for a beneficial mutation to be a back-mutation $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ (see Material and Methods).
    We estimated $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ across all populations sampled and obtained a proportion that varies between 10 and 26\% (table S2).

    The proportion of back-mutations among all beneficial one did not show a clear relationship with the synonymous diversity within a population (Watterson $\theta_w$) (fig.~\ref{fig:diversity}B).
    Indeed, the relationship between $\Ne$ and the proportion of beneficial back-mutations is not trivial and should highly depend on time scales\cite{charlesworth_other_2007}.
    A population that experienced a high $\Ne$ along its evolutionary time should be closer to the optimal state under a stable fitness landscape because of the higher efficacy of selection.
    This should decrease the probability of beneficial back-mutations.
    On the other hand, a high $\Ne$ over a short evolutionary time will induce more extreme fitness effects and a larger proportion of mutations will be beneficial (otherwise effectively neutral), thus increasing the proportion of beneficial back-mutations.
    Overall, the proportion of beneficial back-mutations is expected to be more dependent on effective population size contraction or expansion rather than on the short term effective population size captured by the synonymous diversity\cite{charlesworth_other_2007}.

@@

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.diversity}
        \caption{
            Panel A. For each population, the proportion of beneficial ($\PpolyAdv$), nearly-neutral ($\PpolyNeutral$) and of deleterious ($\PpolyDel$) mutations estimated at the population scale using all SNPs available.
            Population are sorted for increased synonymous diversity.
            Panel B. Proportion of beneficial back-mutations among all beneficial mutations ($\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ in y-axis), shown as a function of the synonymous diversity (Watterson $\theta_w$ in x-axis).
        }
        \label{fig:diversity}
    \end{figure*}

@@

    Protein-coding DNA sequences alignments in placental mammals and their corresponding gene trees were extracted from the \href{https://www.orthomam.univ-montp2.fr}{OrthoMaM} database.
    It contained a total of $116$ mammalian reference sequences in v10c\cite{ranwez_orthomam_2007, douzery_orthomam_2014, scornavacca_orthomam_2019}.
    Genes located on the X, Y chromosomes and the mitochondrial genome were discarded from the analysis, because the level of polymorphism, which is necessary in the population-based analyses, is expected to be different in these three genomic regions.
    Sequences of species for which we used population-level polymorphism (see below), as well as their sister species, were removed from the analysis to ensure independence between the data used in the phylogenetic and population scale.
    Altogether, we analyzed $14,509$ protein-coding DNA sequences for at most $87$ sequences of placental mammals.

>>>>

    Adaptation is considered as the main process driving the diversity of forms and functions across the tree of life, as it enables species to access new ecological niches and also respond to changes in their environment~\cite{darwin_origin_1859}.
    In this context of adaptative evolution, a change in environment is echoed as a change of species' forms and functions, ultimately leaving signatures of accelerated changes in the species' genome~\cite{merrell_adaptive_1994}.
    This signature of an accelerated rate of changes, generating an excess of beneficial mutations is used to quantify adaptation (fig.~\ref{fig:fitness-landscape}A), for specific genes along a specific lineage~\cite{mcdonald_adaptative_1991, smith_adaptive_2002a, welch_estimating_2006}.
    The current availability of large scale genomic data and the development of theoretical models allowed to detect and quantify adaption across genes and lineages, which is now a common practice in evolutionary biology~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    Although these approaches led to a better understanding of the processes shaping molecular evolution, it had the effect to merely associate beneficial mutations with adaptive evolution, which should be dissociated since adaptive evolution is not the only process which can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    Indeed, even in a constant environment for which the population is already adapted, a deleterious mutation in an individual can reach fixation in the population by genetic drift~\cite{Ohta1992}.
    Subsequently, a mutation restoring the ancestral state will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred~\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    More precisely, this restoration of the ancestral state can happen at another loci, in which case it is referred as compensating mutations~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    Conversely, the restoration of the ancestral state can happen at the same locus, in which case it is referred as a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}, which are the focus of this study.
    Taken together, not all beneficial mutations contribute to genetic innovation, and truly adaptive evolution can only be well estimated under the background of beneficial back-mutations~\cite{keightley_what_2010, rice_evolutionarily_2015}.
    It is thus essential to distinguish the underlying reason for the advantage brought by a beneficial mutation, a back-mutation repairing damaged genomes or creating an adaptive innovations.

    For an individual carrying a beneficial mutation, whether this mutation is new innovation as response to a change in environment or a beneficial back-mutation is indistinguishable.
    Likewise, at the level of the population, both will result in a positive transmission bias of the beneficial allele.
    However, their consequences at the macro-evolutionary scale are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promotes phenotype stability and preserves well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Fundamentally, adaptive evolution is unpredictable because it caused by an unpredictable change in the environment and the underlying fitness landscape~\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known.
    As a result, a way to quantify the amount of back-mutation is to estimate the underlying fitness landscape, assuming it is constant and stable across time.

    In this study, fitness landscapes of protein-coding genes are estimated from multi-species sequence alignments.
    Multi-species sequence alignments can be used as a proxy for fitness, since for example a deleterious amino acid will be more likely to be absent in the alignment.
    One of the most used methods to make these predictions is SIFT (Sort Intolerant From Tolerant), which reliably detects deleterious mutations in many taxa~\cite{ng_sift_2003, vaser_sift_2016}.
    However, SIFT scores is not interpretable as a selection coefficient.
    Second, we know that a lot of mechanisms can influence amino acid frequency other than fitness (phylogenetic inertia, mutation biases, GC-biased gene conversion, mutational paths) which are not accounted for in SIFT\@.
    Contrarily, so-called mutation-selection codon models are rooted within a population-genetic formalism~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    These phylogenetic codon models can disentangle processes of mutation at the DNA level and selection at the amino acid level.
    Such codon models provide a nearly-neutral framework of evolution by estimating the fitness landscape over amino acid sequences for each site of the sequence (fig~\ref{fig:method}A) while taking into account non-adaptive processes impacting codon frequency~\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022a}.
    From the fitness landscape obtained by fitting the model to the data, one can thus compute the scaled selection coefficient, denoted $\Sphy$, of any non-synonymous mutations.
    Importantly, mutation-selection codon models assume that the underlying fitness landscape is stable along the phylogeny (as in fig.~\ref{fig:fitness-landscape}B), hence the null subscript $0$ for $\Sphy$.
    As a result, a mutation with a positive $\Sphy$ would be driving the current site toward a fitter amino-acid, and thus considered a beneficial back-mutation repairing existing functions.
    Altogether, our expectation if that beneficial back-mutations are detectable and scattered across the genomes, while at the same time some slightly deleterious mutations are reaching fixations at other loci, generating a balance at which genomes are constantly both damaged and repaired at different loci.
    Even though each of these back-mutations is expected to have a small beneficial effect on their bearer, and that our assumptions may not always be valid, we expect that the genome-wide signature of beneficial back-mutations can be detected and quantified.

    By integrating datasets at different scales (phylogeny and population), we can predict whether substitutions in terminal lineages or any mutation currently segregating in a population are beneficial back-mutations, i.e with a positive $\Sphy$ (fig~\ref{fig:method}A\&B).
    Such a dataset can allow us to address different questions regarding the selection affecting protein-coding DNA sequences.
    Are we empirically observing beneficial back-mutations that reached fixation recently, and one that are currently segregating in populations?
    Are they effectively beneficial for the individual carrying it?
    Conversely, what is the probability for a beneficial mutation to be a back-mutation?
    In other words, can we find empirical evidence and can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    To answer these questions, we estimated the fitness landscape of amino-acids for long-term evolution using available multiple sequence alignments between mammalian species~\cite{ranwez_orthomam_2007, howe_ensembl_2021}.
    We then integrated data at the phylogenetic and population scale to estimate the amount of beneficial mutations across the entire exome (fig.~\ref{fig:method}A-F), for 28 populations from 6 genera (\textit{Equus},  \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}).

@@

    \subsection*{Back-mutations in the terminal lineages}
    We first probed for empirical evidence of beneficial back-mutation in the terminal lineages, which should be effectively beneficial.
    We thus reconstructed the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the mammals' phylogeny (fig.~\ref{fig:method}A).
    We showed that among all the substitutions in the terminal lineages, between 10 and 13\% had $\Sphy > 1$ (table~S1) and between 25 and 30\% had $\Sphy > 0$ (table~S2).
    Substitutions with $\Sphy > 1$ have been beneficial since the ratio of non-synonymous over synonymous divergence, called $\dnds$, was estimated to be between 1.15 and 1.68 in the different lineages (table~S1).
    This meant that beneficial back-mutations with $\Sphy > 1$ reach fixation more frequently than synonymous mutations, which are supposed to be neutral, thus are effectively beneficial.
    This also means that the ratio $\dnds$ usually used to detect adaptation while relying on substitutions between close species~\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale~\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is overinflated due to beneficial back-mutations.
    Then, $\dnds$ computed exclusively on non-synonymous sites with a $\Sphy < 0$ only represent the rate of evolution discarding beneficial back-mutations by construction.
    Thus, if we divide the $\dnds$ estimated on all substitutions by the one estimated only on substitutions predicted to be deleterious, we directly get the proportion of the rate of evolution which is contributed by beneficial back-mutations (see methods).
    This way, we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations (table~S2).

    \subsection*{Back-mutations in the populations}
    Given that beneficial back-mutations are effectively selected for, we then ask which proportion of new mutations are beneficial back-mutations and are also beneficial at the population scale (fig.~\ref{fig:method}C-E).
    For each population with available polymorphism data, we first reconstructed the ancestral sequence (instead of the reference sequence) at the base of the population genealogy.
    From this reference sequence, we computed the scaled selection coefficient ($\Sphy$) for each possible non-synonymous mutations.
    We thus obtained the distribution of fitness effects (DFE) of new mutations from the ancestral sequence in \textit{Homo sapiens} (fig.~\ref{fig:homo-afr-results}A), under the assumption of the stable fitness landscape.
    Because $\Sphy$ is based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ are considered beneficial back-mutations.
    Beneficial back-mutations only represents ~1.5\% of all possible new mutations (red in fig.~\ref{fig:homo-afr-results}A, table~S3).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents ~8.5\% of all observed mutations (red in fig.~\ref{fig:homo-afr-results}B, table~S3).

    Additionally, in populations, the frequencies at which SNPs are segregating is informative of their selective effect, e.g. deleterious SNPs are segregating at lower frequencies because they are purified.
    As such, methods have been developed to estimate the distribution of fitness effect (DFE) at the population scale by gathering information across many SNPs~\cite{eyre-walker_distribution_2006a, eyre-walker_estimating_2009a, galtier_adaptive_2016, tataru_inference_2017}.
    These methods offer a unique opportunity to compare the DFE at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    In order to estimate the DFE at the population level and gather robust statistical estimates, we grouped SNPs observed at the population level (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}) in 5 classes of predicted selection coefficient: severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).
    In each of theses classes, the number of derived alleles in the population for each SNPs is used to generate a site-frequency spectrum (SFS) (fig.~\ref{fig:homo-afr-results}C).
    Compared to synonymous mutations which are supposedly neutral (black line in fig.~\ref{fig:homo-afr-results}C), mutations predicted to be deleterious (blue and green lines in fig.~\ref{fig:homo-afr-results}C) were, as expected, underrepresented at high frequencies in the population, a signature of ongoing purifying selection.
    However, these observations remain qualitative.

    To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE~\cite{tataru_inference_2017, tataru_polydfe_2020}.
    This model uses the count of derived alleles for each frequency class of the SFS to infer the distribution of fitness effects (DFE) of the region under study.
    The model uses the number of sites ($L$) for each category (fig.~\ref{fig:homo-afr-results}A, see Material and Methods), the counts of synonymous derived allele as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral state.
    We could estimate the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogeny of mammals were effectively purged at the population scale.
    Purifying selection is therefore largely predictable and amino acids with low fitness across mammals are effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the scale of the phylogeny of mammals are indeed beneficial for individuals currently bearing them.
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange in fig.~\ref{fig:homo-afr-results}) were effectively composed of neutral, beneficial and deleterious mutations.

    We replicated our experiment across 28 populations of different mammal species (fig.~\ref{fig:heatmap}) and found that SNPs predicted to be deleterious were indeed purged at the population scale (blue and green rows in fig.~\ref{fig:heatmap}A).
    However, SNPs predicted to be weakly beneficial ($\divWeakAdv$) were composed of a mix of neutral and selected mutations with varying proportions across the different populations (orange row in fig.~\ref{fig:heatmap}B).
    Finally, SNPs predicted to be beneficial back-mutations ($\divAdv$) were found to be beneficial with varying proportions across populations (red row in fig.~\ref{fig:heatmap}C).
    The variable proportions found between populations can be explained by the effective number of individuals in the population ($\Ne$), which is well-known to be a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, we used the Watterson $\theta_w$ as a proxy to order populations by their levels of synonymous diversity.
    Higher synonymous diversity is associated with a smaller proportion of nearly-neutral mutations (fig.~\ref{fig:diversity}A\&B).
    This result confirms the nearly-neutral theory and suggests that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) are more likely to discriminate between beneficial and deleterious mutations (fig~S1-S9), while mutations in populations with low diversity (e.g.~\textit{Homo}) are effectively nearly-neutral.

@@

    Finally, we propose an estimator for the probability of a beneficial mutation to be a beneficial back-mutation, i.e. $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ (fig.~\ref{fig:method}C-F).
    We obtained from the analyses above the following probabilities $\mathbb{P} [ \Sphy > 1 ]$, $\mathbb{P} [ \Spop > 1 ]$, $\mathbb{P} [ \Spop > 1  \given  \Sphy > 1]$.
    First, $\mathbb{P} [ \Sphy > 1 ]$ is the probability for a beneficial back-mutation ($\Sphy > 1$) from the ancestral sequence (red fill of histogram in fig~\ref{fig:homo-afr-results}A and table~S3).
    Second, $\mathbb{P} [ \Spop > 1 ]$ is the probability of mutation to be beneficial at the level of the population ($\Spop > 1$), estimated from all available SNPs (black fill in fig~\ref{fig:diversity}A and table~S3).
    Finally, $\mathbb{P} [ \Spop > 1  \given  \Sphy > 1]$ is the probability for a back-mutation ($\Sphy > 1$) to be effectively beneficial at the level of the population ($\Spop > 1$), when we restricted our analysis on SNPs that are already predicted to be beneficial back-mutations (black fill for the category $\Sphy > 1$ in fig~\ref{fig:homo-afr-results}D and table~S3).
    Combining these probabilities using Bayes theorem, we now obtain the probability for a beneficial mutation to be a back-mutation $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ (see Material and Methods).
    We estimated $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ across all populations sampled and obtained a proportion that varies between 10 and 26\% (table~S3).

    The proportion of back-mutations among all beneficial one did not show a clear relationship with the synonymous diversity within a population (Watterson $\theta_w$) (fig.~\ref{fig:diversity}C).
    Indeed, the relationship between $\Ne$ and the proportion of beneficial back-mutations is not trivial and should highly depend on time scales~\cite{charlesworth_other_2007}.
    A population that experienced a high $\Ne$ along its evolutionary time should be closer to the optimal state under a stable fitness landscape because of the higher efficacy of selection.
    This should decrease the probability of beneficial back-mutations.
    On the other hand, a high $\Ne$ over a short evolutionary time will induce more extreme fitness effects and a larger proportion of mutations will be beneficial (otherwise effectively neutral), thus increasing the proportion of beneficial back-mutations.
    Overall, the proportion of beneficial back-mutations is expected to be more dependent on effective population size contraction or expansion rather than on the short term effective population size captured by the synonymous diversity~\cite{charlesworth_other_2007}.

@@

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.diversity}
        \caption{
            Panel A. For each population, the proportion of beneficial ($\PpolyAdv$), nearly-neutral ($\PpolyNeutral$) and of deleterious ($\PpolyDel$) mutations estimated at the population scale using all SNPs available.
            Population are sorted for increased synonymous diversity.
            Panel B. Proportion of nearly-neutral mutations ($\mathbb{P} [ -1 < \Spop < 1]$ in y-axis), shown as a function of the synonymous diversity (Watterson $\theta_w$ in x-axis).
            Panel C. Proportion of beneficial back-mutations among all beneficial mutations ($\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ in y-axis), shown as a function of the synonymous diversity (Watterson $\theta_w$ in x-axis).
        }
        \label{fig:diversity}
    \end{figure*}

@@

    Protein-coding DNA sequences alignments in placental mammals and their corresponding gene trees were extracted from the \href{https://www.orthomam.univ-montp2.fr}{OrthoMaM} database.
    It contained a total of $116$ mammalian reference sequences in v10c~\cite{ranwez_orthomam_2007, douzery_orthomam_2014, scornavacca_orthomam_2019}.
    Genes located on the X, Y chromosomes and the mitochondrial genome were discarded from the analysis, because the level of polymorphism, which is necessary in the population-based analyses, is expected to be different in these three genomic regions.
    Sequences of species for which we used population-level polymorphism (see below), as well as their sister species, were removed from the analysis to ensure independence between the data used in the phylogenetic and population scale.
    Altogether, we analyzed $14,509$ protein-coding DNA sequences for at most $87$ sequences of placental mammals.
